### 2017-02-09 Manuscript/SW_early_life_telomeres.Rmd
Edits

There was positive correlation between RTL measured from different samples taken at different time points during an individual’s life (Fig. 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`). Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL - measured across two samples taken from the same individuals over time - increased with age in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL~*individual*~ measurements (Fig. 2A). To test whether increases in telomere length in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of the same samples to the variance observed among different samples of the same individual. We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 2B). Splitting the longitudinal data into instances of decreasing (i.e. $\Delta$RTL < 0) and increasing (i.e. $\Delta$RTL > 0) telomere length revealed that not only did we observe significantly greater decrease in RTL within individuals compared to within samples (Wilcoxon test: P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also a significantly greater increase (P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig. 2B).

>>>>

There was positive correlation between RTL measured from different samples taken at different time points during an individual’s life (Fig. 2A), but this was very weak (marginal R^2^ = `r round2(r.squaredGLMM(RTL_RTL1)['R2m'])`), and not significant (estimate = `r round3(fixef(RTL_RTL1)['RTL'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_RTL1)['RTL',],lessthan = F),collapse = ', ')`). Although both cross-sectional and longitudinal data indicated a general trend of telomere shortening with age, we found that RTL - measured across two samples taken from the same individuals over time - increased with age in `r round((nrow(subset(dd3,DeltaRTL>0))/nrow(dd3))*100,0)`% of our `r nrow(dd3)` $\Delta$RTL~*individual*~ measurements (Fig. 2A). To test whether increases in telomere length in our dataset could be explained by measurement error, we compared variance in telomere length among repeat measurements of the same samples to the variance observed among different samples of the same individual. We found significantly higher variance in telomere length over individual lifetimes compared to among sample replicates (Levene's test: F = `r round2(leveneTest(DeltaRTL~Group,data=ddL)['F value'][1,1])`; P `r round3(leveneTest(DeltaRTL~Group,data=ddL)['Pr(>F)'][1,1])`; Fig. 2B). Splitting the longitudinal data into instances of decreasing (i.e. $\Delta$RTL < 0) and increasing (i.e. $\Delta$RTL > 0) telomere length revealed that not only did we observe significantly greater decrease in RTL within individuals compared to within samples (Wilcoxon test: P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL<0))$p.value)`), but also a significantly greater increase (P `r round3(wilcox.test(DeltaRTL~Group,data=subset(ddL,DeltaRTL>0))$p.value)`; Fig. 2B). To better understand how longitudinal telomere dynamics vary with age, we exmained patterns of short-term telomere change, including only pairs of samples taken within two-years of each other. We found that tongitudinal telomere length increased with log age both when considered as the absolute degree of change (LMM with $\Delta$RTL~*individual*~ as response; estimate = `r round3(fixef(DeltaRTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`; Fig. 2C), and the likelihood of lengthening (GLMM with lengthened yes/no as binomial response; estimate = `r round3(fixef(DeltaRTLF_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(DeltaRTLF_LogAge)['LogAge',],lessthan = F),collapse = ', ')`; Fig. 2D). Increases in telomere length were most likely to be obsevrved shortly after the juvenile period, at around four years of age, and later in life (although sample sizes for older birds are much smaller; Fig. 2C,D).

### 2017-02-09 Manuscript/SW_early_life_telomeres_tables_and_figures.Rmd
Edits

Fig2A <- ggplot(dd3,
       aes(x = RTL,
           y = RTL1)) +
  ylab(expression('RTL ('*italic(t)*' + 1)')) + 
  xlab(expression('RTL ('*italic(t)*')')) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = c('none'),
        legend.title = element_text(size = 12),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black')) +
    geom_point(col = 'grey')+
  geom_abline(slope = 1,intercept = 0,lty = 2) +
  annotate('text',x = 0.3,y = 1.7,label = 'A')

@@

temp <- subset(dd3,TimeDiff < 4)
temp$MidAge <- (temp$Agemonths+temp$Agemonths1)/2
temp$Gainer <- ifelse(temp$DeltaRTL>0,1,0)

Fig2C <- ggplot(temp, aes(x = Agemonths,
           y = Gainer)) +
  ylab("Proportion gainers") + 
  xlab("Age") +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = c('none'),
        legend.title = element_text(size = 12),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black')) +
  geom_point(col = 'grey')+
  geom_count(alpha = 0.2)+
  geom_smooth()+
  annotate('text',x = -0.2,y = 1,label = 'C')

pdf('Figures/Figure 2.pdf',
    width = 7,
    height = 7,
    paper = 'a4')
multiplot(Fig2A,Fig2B,Fig2C,layout = matrix(c(1,2,3,3),nrow = 2,byrow = T))
dev.off()

>>>>

temp <- subset(dd3,TimeDiff < 3.5)

Fig2A <- ggplot(dd3,
       aes(x = RTL,
           y = RTL1)) +
  ylab(expression('RTL ('*italic(t)*' + 1)')) + 
  xlab(expression('RTL ('*italic(t)*')')) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = c('none'),
        legend.title = element_text(size = 12),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black')) +
    geom_point(col = "grey")+
  geom_abline(slope = 1,intercept = 0,lty = 2) +
  annotate('text',x = 0.3,y = 1.7,label = 'A')

@@

Fig2C <- ggplot(temp, aes(x = Agemonths,
           y = DeltaRTL)) +
  ylab(expression(Delta*'RTL'))+
  xlab("Age (years)") +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = c('none'),
        legend.title = element_text(size = 12),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black')) +
  geom_point(col = "grey")+
  geom_smooth(col = 'black',fill = 'black',method = "lm",formula = y~log(x))+
  geom_abline(slope = 0,intercept = 0,lty = 2)+
  annotate('text',x = -0.2,y = 1,label = 'C')

Fig2D <- ddply(temp,
               .(Agemonths),
               summarise,
               PropGainer = mean(DeltaRTLF),
               n = length(DeltaRTLF)) %>%
ggplot(aes(x = Agemonths,
           y = PropGainer)) +
  ylab("Proportion gainers") + 
  xlab("Age (years)") +
  theme_classic() +
  theme(axis.text.x = element_text(size = 12), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 14,vjust=0.8),
        axis.title.x = element_text(size = 14),
        legend.position = c('none'),
        legend.title = element_text(size = 12),
        axis.line.x = element_line(colour = 'black'),
        axis.line.y = element_line(colour = 'black')) +
  geom_line(col="darkgrey")+
  geom_count(data = temp,aes(y = DeltaRTLF,x = Agemonths),col = "grey")+
  annotate('text',x = 0,y = 1.1,label = 'D')

pdf('Figures/Figure 2.pdf',
    width = 7,
    height = 7,
    paper = 'a4')
multiplot(Fig2A,Fig2B,Fig2C,Fig2D,layout = matrix(c(1,2,3,4),nrow = 2,byrow = T))
dev.off()
